The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-09-26 article.tex
Add @Giappo comments

Current phylogenetic tools use tree priors 
that assume speciation events are independent, 
whilst we know multiple examples of adaptive radiations 
in multiple species families can co-occur, 
for example \richel{@gio: please add examples}.

@@

\richel{explain model here, example is below in the comments}
% In this model, a branching event does not give rise to a new species, but to
% a new species-to-be, called an incipient species. Such an incipient
% species may go extinct, finish its speciation to become a good species, or give
% rise to new incipient species. Protracted speciation may explain observed 
% declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).

Unfortunately, a tree prior according to this model, 
providing the probability of a species tree under the MBD model, 
is unavailable in current Bayesian phylogenetic tools. 
Whilst a likelihood equation has been derived (\richel{cite yourself here}),
it has not been implemented as tree prior yet. 
There are various reasons for this.
First, the computation of the MBD likelihood involves solving a set of 
non-linear differential equations, and while this computation is quite fast, 
it still takes much more time than the corresponding probability 
of the BD model which is a simple analytical formula. 
In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
and hence the total computation will take considerably longer with a PBD tree prior.

@@

The MBD model has \richel{@gio: how many?} parameters, depicted in table \ref{table:parameters}.

\richel{@gio: describe parameter values used here, example is below}
% The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. This means that there is a $0.15 dt$ ($0.35 dt$, $1.0 dt$, $10^9 dt$) probability of speciation completion occurring in an infinitesimal time $dt$. 
% We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
% For each combination of $\lambda$, $\mu$ and $n$, we use a speciation initiation rate $b = b_i = b_g$ that gives a given value of the expected number of species. We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa and the corresponding speciation initiation rates for all parameter combinations are shown in table \ref{table:pbd_parameters}.
% Our parameters are inspired by existing work on protracted speciation \cite{etienne_and_rosindell_2012}\cite{etienne2014}.

@@

We simulate protracted birth-death trees, using the MBD package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of \richel{@gio: parameter values here}, 
we generate incipient species trees with a crown age of 15 million years.
Only trees with the desired number of good taxa are kept.

>>>>

%Current phylogenetic tools use tree priors 
%that assume speciation events are independent, 
%whilst we know multiple examples of adaptive radiations 
%in multiple species families can co-occur, 
%for example 
\richel{@gio: please add examples}.\gio{I've taken the freedom to change your proposed structure. Feel free to change whatever you don't like it. I might be not completely accurate on the biological background.}

Current phylogenetic tools use tree priors that assume that only a single speciation event can occur at the same time.
While this assumption has been proved to be useful to construct a wide variety of successful models, MBD model relaxes this
hypothesis allowing for a description of a different kind of events, where the environmental changes can act as species pump.
This kind of feature can be particularly efficient to describe those kinds of diversification process 
characterized by a tremendous tempo, 
where a great number of species is produced in relatively short time intervals. 
A prime example of that could be given by Cichlid fish diversification in the African Great Lakes: Malawi, Tanganyika and Victoria.

@@

\richel{explain model here, example is below in the comments}
\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
% In this model, a branching event does not give rise to a new species, but to
% a new species-to-be, called an incipient species. Such an incipient
% species may go extinct, finish its speciation to become a good species, or give
% rise to new incipient species. Protracted speciation may explain observed 
% declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).

Unfortunately, a tree prior according to this model, 
providing the probability of a species tree under the MBD model, 
is unavailable in current Bayesian phylogenetic tools. 
Whilst a likelihood equation has been derived (\richel{cite yourself here}),
it has not been implemented as tree prior yet. 
There are various reasons for this.
First, the computation of the MBD likelihood involves solving a set of 
non-linear differential equations \gio{@richel: are they actually non-linear?}, and while this computation is quite fast, 
it still takes much more time than the corresponding probability 
of the BD model which is a simple analytical formula. 
In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
and hence the total computation will take considerably longer with a PBD tree prior.

@@

The MBD model has 4 parameters, depicted in table \ref{table:parameters}. Parameters $\lambda$ and $\mu$ correspond, respectively, to the usual per-species speciation and extinction rates. The model also introduces two new parameters: $\nu$ is the total rate for an environmental change to trigger, which leads to a potentially multiple speciation event. If such event triggers each species present at that moment in time can undergo a speciation event with probability $q$.

\richel{@gio: describe parameter values used here, example is below}\gio{@richel: I described the meaning of each parameter. I guess it is fine in this way. For the setting I think we have to wait to decide which values are actually worth a try}
% The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. This means that there is a $0.15 dt$ ($0.35 dt$, $1.0 dt$, $10^9 dt$) probability of speciation completion occurring in an infinitesimal time $dt$. 
% We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
% For each combination of $\lambda$, $\mu$ and $n$, we use a speciation initiation rate $b = b_i = b_g$ that gives a given value of the expected number of species. We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa and the corresponding speciation initiation rates for all parameter combinations are shown in table \ref{table:pbd_parameters}.
% Our parameters are inspired by existing work on protracted speciation \cite{etienne_and_rosindell_2012}\cite{etienne2014}.

@@

We simulate protracted birth-death trees, using the MBD package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $(\lambda, \mu, \nu, q) $\richel{@gio: parameter values here}, 
we generate incipient species trees with a crown age of 15 million years \gio{are we sure we wanna try 15 million years. so far i've been trying only with 10 million years, which most of the time is working really well}.
Only trees with the desired number of good taxa are kept.
